The importance of a complete statistical description of dependence between variables for ecology and related fields

### 2019-01-10 Paper.Rmd
firther edits

Although these results do convincingly show non-normal copula structure, 
we caution that the model-averaged tail-dependence statistics may have been biased 
because even the lowest-AIC copula 
(`r BivMS_res_Loecke$InfCritRes$copname[which.max(BivMS_res_Loecke$InfCritRes$AICw)]`) 
was a poor fit ($p=$ `r BivMS_res_Loecke$GofRes_CvM` and `r BivMS_res_Loecke$GofRes_KS` for the 
Cramer-von Mises and Kolmogorov-Smirnov goodness-of-fit tests, respectively, to within the precision 
available from `BiCopGofTest` in the `VineCopula` package). However, nonparametric statistics verified
both that tail dependence was distinct from that of a normal copula for the soil data, and that it 
was asymmetric (Fig. \ref{SM-fig_soilCN_nonparam}, Table \ref{SM-tab_soilCN_asym}). 
The values of the statistics $\cor_{l_b,u_b}$, $\Ps_{l_b,u_b}$ and $\Dsq_{l_b,u_b}$ for real data
were compared to distributions of their values on 1000 Kendall- or Spearman-preserving N surrogates
of the data,
separately in two comparisons for each of the ranges $(l_b,u_b)=(0,0.1),\ldots,(0.9,1)$ 
(Fig. \ref{SM-fig_soilCN_nonparam}). Results show that tail dependence of data is stronger in 
both the lower and upper tails than would be expected under a null hypothesis of a normal copula.
The values of the statistics $\cor_{0,0.1}-\cor_{0.9,1}$, $\Ps_{0,0.1}-\Ps_{0.9,1}$ and
$\Dsq_{0.9,1}-\Dsq_{0,0.1}$ for real data were also compared to distributions of their values
on 1000 Kendall- or Spearman-preserving N surrogates, separately in two comparisons for each
of the $(l_b,u_b)$ ranges listed above (Table \ref{SM-tab_soilCN_asym}). Results showed that
upper-tail dependence was significantly stronger than lower-tail dependence.

@@

We present our analysis of the green spruce aphid abundance data (see Data).
Independence was tested for each pair of the 10 sampling locations, and was 
rejected in every case.
<!--***DAN: Right, Shyamolina? I am assuming that because "Indep" does not 
appear anywhere in Table S11, this is correct, but please confirm or correct-->
Model selection among copulas was carried out separately for each pair of locations.
Best-fitting (lowest-AIC) copulas were non-normal
for the large majority of location pairs (Table \ref{SM-tab_count_bestcop}), and 
AIC weights for the best-fitting copulas (Table \ref{SM-tab_count_AICw}) 
were typically substiantially 
larger than those for the N copula (Table \ref{SM-tab_count_AICw_normal}), answering 
the first part of Q1 from the Introduction for these data.
Goodness of fit tests based on a Cramer-von Mises statistic 
(Table \ref{SM-tab_count_p_CvM}) in every case 
failed to reject the hypothesis that the AIC-best copula family was also 
an objectively adequate description of 
the data, and likewise for goodness of fit tests based on a Kolmogorov-Smirnov 
statistic (Table \ref{SM-tab_count_p_KS}), demonstrating that the collection of 
copula families we used was sufficiently broad that the copulas of all pairs of 
locations could be characterized by at least one of them, for these data.
Model-averaged lower- (Table \ref{SM-tab_count_LTdep_AICw}) and upper-tail
(Table \ref{SM-tab_count_UTdep_AICw}) dependence statistics were exclusively 
positive and were frequently substantially large, thereby differing 
from what an N copula would give (i.e., 0), and answering the second part of 
Q1 for these data. Apparently these aphid data have greater tail dependence 
(both lower and upper) than would be expected from an N-copula null hypothesis. 
The model-averaged lower-tail dependence statistic minus the model-averaged 
upper-tail dependence statistic (Table \ref{SM-tab_count_LTmUT}) was positive 
for all but a few locations pairs (`r sum((RES_aphid_count$LTdep_AICw-RES_aphid_count$UTdep_AICw)>0,na.rm=T)` 
out of `r (dim(RES_aphid_count$gfc_p_CvM)[1]^2-dim(RES_aphid_count$gfc_p_CvM)[1])-2*RES_aphid_count$num_indep_loc_pair`), 
indicating that lower-tail dependence across space was much stronger than upper-tail 
dependence. Thus the spatial synchrony of rarity in green spruce aphid is stronger 
than the spatial synchrony of outbreak events. This also answers the third part of Q1.

<!--Non-parametric analysis of tail dependence, green spruce abundance data-->

As mentioned in \nameref{NPA} and Appendix \ref{SM-spatial_resamp}, mean values of our statistics ($\cor_l$, $\cor_u$, $\Ps_l$, $\Ps_u$, $\Dsq_l$, $\Dsq_u$, 
$\cor_l-\cor_u$, $\Ps_l-\Ps_u$, $\Dsq_u-\Dsq_l$), as well as mean Kendall and Spearman correlations, were 
computed across all pairs of sampling locations, and confidence intervals of these means were computed via a spatial
resampling scheme (Table \ref{SM-tab_count_resamp}). Confidence intervals of
$\cor_l$, $\cor_u$, $\Ps_l$, $\Ps_u$, $\Dsq_l$, and $\Dsq_u$ always excluded zero and mean values were positive,
indicating tail dependence. Confidence intervals of $\cor_l-\cor_u$, $\Ps_l-\Ps_u$, $\Dsq_u-\Dsq_l$ also always 
exluded $0$ and were positive, indicating asymmetry of tail dependence (stronger lower tail dependence than upper), a feature that data with Normal copula structure do not have, providing further evidence answering the third part of question \hyperref[Intro]{\textbf{(Q1)}}. Visual illustration of those results is discussed in Appendix \ref{SM-visual_npa}.

<!--multivariate summary-->
We did model selection and non-parametric analysis with all 20 aphid and 22 plankton species and found our two approaches were consistent to each other. Here, to present our results we chose one species from each category. We summarized our results columnwise in Table \ref{tab_multivar_summary} for all multivariable datasets (i.e., abundance data for green spruce aphid and firstflight date data for leaf-curling plum aphid from `r nrow(RES_aphid_count$gfc_numBS)` sampling locations, abundance data for plankton from UK seas from $14$ sampling locations, 
methane flux data from $13$ sampling locations).

@@

@@

row.names(tab_multivar_summary)<-NULL
library(kableExtra)
library(dplyr)
library(tibble)
dt <- tibble(
            Description = c("1. Total number of location pairs (excluding self-comparison)",
           "2. Number of location pairs having non-independent copula structures",
           "3. Number of non-Normal copula as best fit",
           "4. Average AIC-weights of best fit copula from all location pair",
           "5. Average AIC-weights of Normal copula from all location pair",
           "6. Number of good-fit location pair*",
           "7. 95% CI of model-avg. lower tail-dep. from all location pair",
           "8. 95% CI of model-avg. upper tail-dep. from all location pair",
           "9. Percentage of location pairs showing stronger lower tail dependence",
           "10. 95% CI of (Pl - Pu) from all location pair")
           )

@@

knitr::kable(tab_multivar_summary, 
             format="latex", 
             caption = "Summary of results for multivariate data-sets\\label{tab_multivar_summary}",
             booktabs=T, linesep = "\\addlinespace",
             col.names = c(" ","Green-spruce aphid's count",
                        "Leaf-plum curling aphid's firstflight",
                        "Ceratium furca's count","Methane-flux")
             )%>%column_spec(2:5,width="2.5cm")%>%
             column_spec(1, width = "7cm")%>%kable_styling(latex_options = "basic",position = "center")%>%
   add_footnote("Good fit was considered for best fit copula with p-values > 0.01 (both Cramer-von Mises and Kolmogorov-Smirnov method)", notation="symbol", threeparttable = T, escape = F)
```

>>>>

Although these results do convincingly show non-normal copula structure, 
we caution that the model-averaged tail-dependence statistics may have been biased 
because even the lowest-AIC copula 
(`r BivMS_res_Loecke$InfCritRes$copname[which.max(BivMS_res_Loecke$InfCritRes$AICw)]`) 
was a poor fit ($p=$ `r BivMS_res_Loecke$GofRes_CvM` and `r BivMS_res_Loecke$GofRes_KS` for the 
Cramer-von Mises and Kolmogorov-Smirnov goodness-of-fit tests, respectively, to within the precision 
available from `BiCopGofTest` in the `VineCopula` package). However, nonparametric statistics verified
both that tail dependence was distinct from that of an N copula for the soil data, and that it 
was asymmetric (Fig. \ref{SM-fig_soilCN_nonparam}, Table \ref{SM-tab_soilCN_asym}). 
The values of the statistics $\cor_{l_b,u_b}$, $\Ps_{l_b,u_b}$ and $\Dsq_{l_b,u_b}$ for real data
were compared to distributions of their values on 1000 Kendall- or Spearman-preserving N surrogates
of the data,
separately in two comparisons for each of the ranges $(l_b,u_b)=(0,0.1),\ldots,(0.9,1)$ 
(Fig. \ref{SM-fig_soilCN_nonparam}). Results show that tail dependence of data is stronger in 
both the lower and upper tails than would be expected under a null hypothesis of a normal copula.
The values of the statistics $\cor_{0,0.1}-\cor_{0.9,1}$, $\Ps_{0,0.1}-\Ps_{0.9,1}$ and
$\Dsq_{0.9,1}-\Dsq_{0,0.1}$ for real data were also compared to distributions of their values
on 1000 Kendall- or Spearman-preserving N surrogates, separately in two comparisons for each
of the $(l_b,u_b)$ ranges listed above (Table \ref{SM-tab_soilCN_asym}). Results showed that
upper-tail dependence was significantly stronger than lower-tail dependence.

@@

We present our analysis of the green spruce aphid abundance data (see Data).
Independence was tested for each pair of the 10 sampling locations, and was 
rejected in every case.
<!--***DAN: Right, Shyamolina? I am assuming that because "Indep" does not 
appear anywhere in Table S11, this is correct, but please confirm or correct-->
Model selection among copulas was carried out separately for each pair of locations.
Best-fitting (lowest-AIC) copulas were non-normal
for the large majority of location pairs (Table \ref{SM-tab_count_bestcop}), and 
AIC weights for the best-fitting copulas (Table \ref{SM-tab_count_AICw}) 
were typically substiantially 
larger than those for the N copula (Table \ref{SM-tab_count_AICw_normal}), answering 
the first part of Q1 from the Introduction for these data.
Goodness of fit tests based on a Cramer-von Mises statistic 
(Table \ref{SM-tab_count_p_CvM}) in every case 
failed to reject the hypothesis that the AIC-best copula family was also 
an objectively adequate description of 
the data, and likewise for goodness of fit tests based on a Kolmogorov-Smirnov 
statistic (Table \ref{SM-tab_count_p_KS}), demonstrating that the collection of 
copula families we used was sufficiently broad that the copulas of all pairs of 
locations could be characterized by at least one of them, for these data.
Model-averaged lower- (Table \ref{SM-tab_count_LTdep_AICw}) and upper-tail
(Table \ref{SM-tab_count_UTdep_AICw}) dependence statistics were exclusively 
positive and were frequently substantially large, thereby differing 
from what an N copula would give (i.e., 0), and answering the second part of 
Q1 for these data. Apparently these aphid data have greater tail dependence 
(both lower and upper) than would be expected from an N-copula null hypothesis. 
Model-averaged lower-tail dependence statistics minus model-averaged 
upper-tail dependence statistics (Table \ref{SM-tab_count_LTmUT}) were positive 
for all but a few locations pairs (`r sum((RES_aphid_count$LTdep_AICw-RES_aphid_count$UTdep_AICw)>0,na.rm=T)` 
out of `r (dim(RES_aphid_count$gfc_p_CvM)[1]^2-dim(RES_aphid_count$gfc_p_CvM)[1])-2*RES_aphid_count$num_indep_loc_pair`), 
indicating that lower-tail dependence across space was much stronger than upper-tail 
dependence. Thus the spatial synchrony of rarity in green spruce aphid is stronger 
than the spatial synchrony of outbreak events. This also answers the third part of Q1.

<!--Non-parametric analysis of tail dependence, green spruce aphid abundance data-->
Nonparametric statistics verified both that tail dependence was distinct from that 
of an N copula for the green spruce aphid abundance data, and that it was asymmetric.
Mean values across all pairs of locations of the statistics 
$\cor_l$, $\cor_u$, $\Ps_l$, $\Ps_u$, $\Dsq_l$, and $\Dsq_u$ were positive and
confidence intervals excluded $0$ (the value for an N copula), confirming tail 
dependence distinct from that of an N copula (Table \ref{SM-tab_count_resamp}). 
Mean values of the statistics $\cor_l-\cor_u$, $\Ps_l-\Ps_u$, and $\Dsq_u-\Dsq_l$
also were always positive and had confidence intervals that excluded $0$ 
(Table \ref{SM-tab_count_resamp}), confirming asymmetry of tail dependence and 
stronger lower- than upper-tail dependence. Each of the statistics 
was plotted against geographic (great-circle) distances between sampling locations, 
across all pairs of sampling locations (Fig. \ref{SM-fig_aphid_count}).

<!--multivariate summary-->
Table \ref{tab_multivar_summary} summarizes some of our results using our
multivariate data. In addition to summaries of the green spruce aphid abundance data results 
elaborated in detail above, the table also summarizes results of analyzing 
leaf-curling plum aphid first flight data, abundance data for the dinoflagellate
\textit{Ceratium furca}, and methan flux data (see Data).

We did model selection and non-parametric analysis with all 20 aphid and 22 plankton species and found our two approaches were consistent to each other. Here, to present our results we chose one species from each category.

@@

We also carried out the same analyses for abundance and first-flight data for the 18 
aphids species for which we had data other than the green spruce and leaf-curling plum aphids,
as well as for the 21 plankton taxa for which we had abundance data other 
\textit{Ceratium furca} (results not shown). The conclusions were generally supported that 
non-N copula structure and tail dependence, as well as asymmetry of tail dependence, are 
very common.

@@

row.names(tab_multivar_summary)<-NULL
library(kableExtra)
library(dplyr)
library(tibble)
dt <- tibble(
            Description = c("1. Location pairs (excluding self comparisons)",
           "2. Number of non-independent pairs",
           "3. Number of pairs with non-N copula as best fit",
           "4. Average AIC-weights of best-fitting copulas across location pairs",
           "5. Average AIC-weights of N copula across pairs",
           "6. Number of well-fitting location pairs",
           "7. 2.5th and 97.5 quantiles of model-avg. lower-tail dependence",
           "8. 2.5th and 97.5th quantiles of model-avg. upper-tail dependence",
           "9. Percent pairs showing stronger lower tail dependence",
           "10. 95% CI of average P_l - P_u across all location pair")
           )

@@

knitr::kable(tab_multivar_summary, 
             format="latex", 
             short.caption = "Summary of results for multivariate data sets"
             caption = "Summary of results for multivariate data-sets. Rows 3-9 of the table were computed for the non-independent pairs (row 2) only; row 10 was for all location pairs. A well-fitting location pair (row 6) was one for which the best-fitting copula had $p$-values $>0.01$ for both the Cramer-von Mises and Kolmogorov-Smirnov goodness of fit tests. \\label{tab_multivar_summary}",
             booktabs=T, linesep = "\\addlinespace",
             col.names = c(" ","Green-spruce aphid's count",
                        "Leaf-plum curling aphid's firstflight",
                        "Ceratium furca's count","Methane-flux")
             )%>%column_spec(2:5,width="2.5cm")%>%
             column_spec(1, width = "7cm")%>%kable_styling(latex_options = "basic",position = "center")
```
